Supplementary methods

### 2020-08-24 text/supplemental_methods.Rmd

@@

Using the model described above we fit the data using Stan via the rstan package
in R. Posterior inferences are shown in **Figure 7**.

**supplementary figure S5** gives a comparison on log-scale between
the post-fit survival curves and the CDC data it is based on.

@@

>>>>

---
title: "Supplementary methods"
author: "Peter T. Nguyen, Simon G. Coetzee, Daniel L. Lakeland, and Dennis J. Hazelett"
date: "8/24/2020"
bibliography: references.bib
output:
  word_document:
    reference_docx: "../text/word-style-reference-ttmanu.docx"
always_allow_html: TRUE
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
library(foreach)
library(sqldf)
library(tidyverse)
library(ggthemes)
library(data.table)
library(biomaRt)
library(dplyr)
library(GenomicRanges)
library(SummarizedExperiment)
library(reactome.db)
library(plyr)
library(cluster)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
library(dbscan)
library(class)
library(TCGAbiolinks)
library(patchwork)
library(ggplotify)
library(ComplexHeatmap)
library(knitr)
library(kableExtra)
library(UpSetR)
library(cowplot)
library(doRNG)
library(uwot)
library(janitor)
library(FactoMineR)
library(seriation)
library(rtracklayer)
library(poibin)
library(flextable)
library(plot3D)
library(ggfortify)
#library(maftools)
library(survival)
library(ggalluvial)
```

## Supplementary Methods

### Clustering

In order to classify tumors using this dataset, we used multiple correspondence
analysis (MCA) [@JSSv025i01]. First, we determined the number of dimensions
containing useful information by selecting the eigenvalue with the most
explanatory power, using the average of 100 permutations of the data as baseline
(**figure SM1A**). We then chose the maximum eigenvalue for which the p-value
remained ≤ 0.05 (see cutoff in **figure SM1B**). Then we performed a UMAP
analysis, both in order to summarize the MCA graphically, and as a preprocessing
step to boost the performance of density based clustering. The resulting map was
notable for its lobed structure, with several reproducible projections
regardless of random seed setting. A representative version of this 3D UMAP is
shown in **figure SM1C**, rotated to enhance the visibility of the major features.
Following this spatial mapping we attempted to define groupings of similar
tumors within the spatial map using HDBSCAN, which performs hierarchical
clustering and provides metrics of cluster stability and probabilities of
cluster membership for each node. However, HDBSCAN is sensitive to several
parameters; key for our analysis are the minimum number of tumor samples in a
cluster that capture the maximum number of tumor samples, measured by
probability of membership of ≥ 5% in at least one class. Thus, we created a
score metric as the fraction of classified tumors with max probability < 5% in
one class and chose a cluster size of 92 to minimize the score function
(**figure SM1D**). HDBSCAN with these settings resulted in ten distinct
high-density clusters which we then projected onto the UMAP (**figure SM1E**).
This classified 6,038 out of 7,607 tumors but still resulted in a significant
fraction of unclassified tumor samples. Since we ultimately wish to be able to
classify any tumor using this scheme, we performed k Nearest Neighbors (kNN)
analysis, which computes a similarity metric to every tumor in the set and then
lets the _k_ most similar tumors "vote" as to the identity of the query tumor
sample based on their classification labels. We set _k_ to be the square root of
the number of tumor samples (87). Using this method, we assigned cluster
membership to the remaining tumors (**figure SM1F**).

```{r, results='asis'}
cat("\\newpage")
```

#### Figure SM1
```{r child='../code/figureSM1.Rmd'}
```

**Figure SM1: Method for tumor classification using pathway aberration data** 
a) MCA analysis was performed to determine the eigenvalue with the most
explanatory power from the categorical data. Average of 100 permutations of the
data was used as a baseline. b) The p-value of each eigenvalue was plotted and
the max eigenvalue with a pvalue < 0.05 was selected. c) The resulting MCA was
projected down to 3 dimensions for the purpose of density based clustering (only
the first 2 dimensions are shown). Each dot represents a tumor d) HDBSCAN was
used to perform hierarchical density based clustering tuned to minimize the
fraction of unassigned tumors. e) HDBSCAN clustered tumors were visualized on
the UMAP projection; tumors are colored by cluster membership (see legend). This
results in some tumors remaining unclassified (grey points). f) kNN was used to
assign membership to the remaining tumors.

### Survival

@@

**supplementary figure SM2** gives a comparison on log-scale between
the post-fit survival curves and the CDC data it is based on.

@@

### References

### 2020-08-24 text/survival.Rmd

---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. Kaplan-Meier plots reveal large differences between
cancer types as expected, and some modest differences between our classes (not
shown). These analyses are profoundly limited by confounding factors of age and
stage at diagnosis, sex, ethnicity, and tissue-specific disease progression, to
name the obvious ones. To explore these ideas, we tested models of survival
using public longevity data from the CDC, accounting for age, gender and
ethnicity. Our model assumes that at birth each person has a small initial
probability per unit time to die, and that this probability grows exponentially
in time at some baseline rate that matches the observed CDC data. The rate of
growth in risk is assumed to jump to a higher level at diagnosis, which we take
as a proxy for onset. This represents an accelerated aging type model. To fit
the model, we constrain the baseline risks using the available CDC life tables.
We then constrain the cancer specific multiplier using the longevity data within
the cancer tumor dataset. Our model estimates the effect of cancer type and
class-specific cancer effects independently, resulting in a cancer and class
specific estimate of the aging rate multiplier.

>>>>

---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are profoundly limited by
confounding factors of age and stage at diagnosis, sex, ethnicity, and
tissue-specific disease progression, to name the obvious ones. To explore these
ideas, we used Bayesian inference to test models of survival using public
longevity data from the CDC, accounting for age, gender and ethnicity. Our model
estimates the effect of cancer type and class-specific cancer effects
independently, resulting in a cancer and class specific estimate of the aging
rate multiplier.
